

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Instructions

```{r code_file_instructions, echo=FALSE, eval=FALSE}
## INSTRUCTIONS:
## 1. TO RUN A COMMAND, PLACE CURSOR INSIDE COMMAND
##    OR HIGHLIGHT COMMAND(S) AND HIT CTRL-ENTER (COMMAND-ENTER FOR MACS)
## 2. TO RUN ALL CODE FROM BEGINNING OF FILE TO CURRENT LINE,
##    HIT CTRL-ALT-B (RSTUDIO ONLY)
## 3. COMMANDS THAT BEGIN WITH # ARE COMMENTS AND WILL NOT BE EXECUTED
## 4. NOTE: A FEW COMMANDS MAY PURPOSEFULLY RESULT IN ERRORS FOR TEACHING PURPOSES
## 5. USE CTRL-F TO FIND
```

## Load Required Libraries

```{r load_libraries}
library(ggplot2)
library(MASS)
```

## Scatter Plot Example

```{r ggplot_scatter, fig.cap="geom_point() inherits x and y aesthetics"}
ggplot(txhousing, aes(x=volume, y=sales)) +
  geom_point()
```

```{r ggplot_no_geom, fig.cap="Without a geom or stat, just axes"}
ggplot(data = txhousing, aes(x=volume, y=sales))
```

```{r override, fig.cap="Both geoms inherit aesthetics from ggplot, but geom_rug() also adds color aesthetic"}
ggplot(txhousing, aes(x=volume, y=sales)) +
  geom_point() +
  geom_rug(aes(color=listings))
```

## Mapping Aesthetic Example

```{r mapping_color, fig.cap="Color of points varies with number of listings"}
ggplot(txhousing, aes(x=volume, y=sales)) +
  geom_point(aes(color=listings))
```

```{r setting_color, fig.cap="Color of points set to constant green"}
ggplot(txhousing, aes(x=volume, y=sales)) +
  geom_point(color="green")
```

```{r color_inside_aes, fig.cap="Aesthetic set to constant within aes() leads to unexpected results"}
ggplot(txhousing, aes(x=volume, y=sales)) +
  geom_point(aes(color="green"))
```

## Histogram and Density Plots

```{r hist, fig.cap="Histograms visualize distribution of variable mapped to x"}
ggplot(txhousing, aes(x=median)) + 
  geom_histogram()
```

```{r density, fig.cap="Density plots visualize smoothed distribution of variable mapped to x"}
#Denisty plots are basically smoothed histograms.

#Density plots, unlike histograms, can be plotted separately by group by mapping #a grouping variable to color
ggplot(txhousing, aes(x=median)) + 
  geom_density()
```

```{r density_color, fig.cap="Densities of median price by month"}
ggplot(txhousing, aes(x=median, color=factor(month))) + 
  geom_density()
```

## Boxplot Example

```{r boxplot, fig.cap="Boxplots are useful to compare distribution of y variable across levels of x variable"}
ggplot(txhousing, aes(x=factor(year), y=median)) + 
  geom_boxplot()
```

## Barplot Example
<!-- Bar plots are often used to display frequencies of factor (categorical) variables. -->

<!-- geom_bar() by default produces a bar plot where the height of the bar represents counts of each x-value. -->

<!-- The color that fills the bars is not controlled by color, but instead by fill, which can only be mapped to a factor (categorical) variable. We can visualize a crosstabulation of variables by mapping one of them to fill in geom_bar(): -->

```{r barplot, fig.cap="geom_bar displays frequencies of levels of x variable"}
ggplot(diamonds, aes(x=cut)) + 
  geom_bar()
```

```{r barplot_fill, fig.cap="Frequencies of cut by clarity"}
ggplot(diamonds, aes(x=cut, fill=clarity)) + 
  geom_bar()
```

## Line Graph Example
<!-- Line graphs depict covariation between variables mapped to x and y with lines instead of points. -->

<!-- geom_line() will treat all data as belonging to one line unless a variable is mapped to one of the following aesthetics to group the data into separate lines: -->

<!-- group: lines will look the same -->
<!-- color: line colors will vary with mapped variable -->
<!-- linetype: line patterns will vary with mapped variable -->
<!-- Let’s first examine a line graph with no grouping: -->

```{r line_graph, fig.cap="Line graph of sales over time, separate lines by city"}
ggplot(txhousing, aes(x=date, y=sales, group=city)) + 
  geom_line()
```

```{r line_graph_color, fig.cap="Line graph of sales over time, colored and grouped by city"}
ggplot(txhousing, aes(x=date, y=sales, color=city)) + 
  geom_line()
```

## Scatter Plots
<!-- Scatter plots depict the covariation between pairs of variables (typically both continuous). -->

<!-- geom_point() depicts covariation between variables mapped to x and y. -->

<!-- Scatter plots are among the most flexible graphs, as variables can be mapped to many aesthetics such as color, shape, size, and alpha. -->

```{r scatter_1, fig.cap="Scatter plot of volume vs sales"}
ggplot(txhousing, aes(x=volume, y=sales)) + 
  geom_point()
```

```{r scatter_2, fig.cap="Scatter plot of volume vs sales, colored by number of listings, transparent by median price, and sized by inventory"}
ggplot(txhousing, aes(x=volume, y=sales, color=listings, alpha=median, size=inventory)) + 
  geom_point()
```


## Line Graph Examples

```{r line_graph_1, fig.cap="Line graph of sales over time, separate lines by city"}

```

```{r line_graph_2, fig.cap="Line graph of sales over time, no grouping results in garbled graph"}

```

```{r line_graph_3, fig.cap="Line graph of sales over time, colored and grouped by city"}

```

## Statistical Summaries
<!-- The stat functions statistically transform data, usually as some form of summary, such as the mean, or standard devation, or a confidence interval. -->

<!-- Each stat function is associated with a default geom, so no geom is required for shapes to be rendered. -->

<!-- stat_summary(), perhaps the most useful of all stat functions, applies a summary function to the variable mapped to y for each value of the x variable. The default summary function is mean_se(), with associated geom geom_pointrange(), which will produce a plot of the mean (dot) and standard error (lines) of the variable mapped to y for each value of the x variable. -->

```{r stat_summary_1, fig.cap="Mean and standard errors of sales by year"}
ggplot(txhousing, aes(x=year, y=sales)) + 
  stat_summary()
```

## Example Scales

```{r example_scales, echo=FALSE}
color_scale <- data.frame(color=c("red", "blue"), treat=c("ozone", "control"))
knitr::kable(color_scale)
```

## Customizing Scales

```{r scale_default, fig.cap="Default color scale"}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) + 
  geom_point()
```
#
#  can use scale_colour_manual() to specify which colors we want to use in values=:
<!-- Scale functions for the axes -->
<!-- Remember that x and y are aesthetics, and the two axes visualize the scale for these aesthetics. -->

<!-- Thus, we use scale functions to control to the scaling of these axes. -->

<!-- When y is mapped to a continuous variable, we will typically use scale_y_continuous() to control its scaling (use scale_y_discrete() if y is mapped to factor). Similar functions exist for the x aesthetic. -->

<!-- A description of some of the important arguments to scale_y_continuous(): -->

<!-- breaks: at what data values along the range of of the axis to place tick marks and labels -->
<!-- labels: what to label the tick marks -->
<!-- name: what to title the axis -->
<!-- Our current graph of volume vs sales has y-axis tick marks at 0, 5000, 10000, and 15000 -->

```{r scale_manual, fig.cap="Using scale_color_manual to respecify colors"}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) + 
  geom_point() +
  scale_color_manual(values=c("red", "yellow", "green", "blue", "violet"))
```

# Let’s put tick marks at all grid lines along the y-axis using the breaks argument of scale_y_continuous:

```{r scale_ticks, fig.cap="Changing y-axis tick marks"}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) + 
  geom_point() +
  scale_y_continuous(breaks=c(0,2500,5000,7500,10000,12500,15000,17500))
```
#
# Now let’s relabel the tick marks to reflect units of thousands (of dollars) using labels:
#

```{r scale_labels, fig.cap="Relabeling y-axis tick marks"}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) + 
  geom_point() +
  scale_y_continuous(breaks=c(0,2500,5000,7500,10000,12500,15000,17500),
                     labels=c(0,2.5,5,7.5,10,12.5,15,17.5))
```

```{r scale_axis_title, fig.cap="New y-axis title"}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) + 
  geom_point() +
  scale_y_continuous(breaks=c(0,2500,5000,7500,10000,12500,15000,17500),
                     labels=c(0,2.5,5,7.5,10,12.5,15,17.5),
                     name="Price (thousands of dollars)")
```
<!-- Although we can use scale functions like scale_x_continuous() to control the limits and titles of the x-axis, we can also use the following shortcut functions: -->

<!-- lims(), xlim(), ylim(): set axis limits -->
<!-- xlab(), ylab(), ggtitle(), labs(): give labels (titles) to x-axis, y-axis, or graph; labs can set labels for all aesthetics and title -->
<!-- To set axis limits, supply a vector of 2 numbers (inside c(), for example) to one of the limits functions: -->



```{r x_axis_limits, fig.cap="Restricting axis limits will zoom in"}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) + 
  geom_point() +
  xlim(c(1,3))
```

## Custom Labels
#We can use labs() to specify an overall titles for the overall graph, the axes,and legends (guides).

```{r custom_labels, fig.cap="Respecifying all titles with labs"}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) + 
  geom_point() +
  labs(x="CARAT", y="PRICE", color="CUT", title="CARAT vs PRICE by CUT")
```

## Guides and Legends
<!-- Guides visualize scales -->
<!-- Guides (axes and legends) visualize a scale, displaying data values and their matching aesthetic values. -->

<!-- the x-axis visualizes the mapping of data values to a position along the x-axis. -->
<!-- a color scale guide (legend) displays the colors mapped to data values. -->
<!-- Most guides are displayed by default. The guides() function sets and removes guides for each scale. -->

<!-- Here we use guides() to remove the color scale legend: -->

```{r remove_legend, fig.cap="Color legend removed"}
ggplot(txhousing, aes(x=volume, y=sales, color=median)) + 
  geom_point() +
  guides(color="none")
```

## Faceting
<!-- Split plots into small multiples (panels) with the faceting functions, facet_wrap() and facet_grid(). The resulting graph shows how each plot varies along the faceting variable(s). -->

<!-- facet_wrap() wraps a ribbon of plots into a multirow panel of plots. Inside facet_wrap(), specify ~, then a list of splitting variables, separated by +. The number of rows and columns can be specified with arguments nrow and ncol. -->

```{r facet_wrap, fig.cap="Carat vs Price, paneled by cut with facet_wrap()"}
ggplot(diamonds, aes(x=carat, y=price)) + 
  geom_point() + 
  facet_wrap(~cut)
```
<!-- facet_grid() allows direct specification of which variables are used to split the data/plots along the rows and columns. Put the row-splitting variable before ~, and the column-splitting variable after. The character . specifies no faceting along that dimension. -->
```{r facet_grid, fig.cap="Carat vs Price, paneled by clarity and cut with facet_grid()"}
ggplot(diamonds, aes(x=carat, y=price)) + 
  geom_point() + 
  facet_grid(clarity~cut)
```


## Line Graph Examples

```{r line_graph_1, fig.cap="Line graph of sales over time, separate lines by city"}

```

```{r line_graph_2, fig.cap="Line graph of sales over time, no grouping results in garbled graph"}

```

```{r line_graph_3, fig.cap="Line graph of sales over time, colored and grouped by city"}

```

## Statistical Summaries

```{r stat_summary_1, fig.cap="Mean and standard errors of sales by year"}

```

## Example Scales

```{r example_scales, echo=FALSE}

```

## Customizing Scales

```{r scale_default, fig.cap="Default color scale"}

```

```{r scale_manual, fig.cap="Using scale_color_manual to respecify colors"}

```

## Themes and Customizations
<!-- Themes control elements of the graph not related to the data. For example: -->

<!-- background color -->
<!-- size of fonts -->
<!-- gridlines -->
<!-- color of labels -->
<!-- To modify these, we use the theme() function, which has a large number of arguments called theme elements, which control various non-data elements of the graph. -->

<!-- Some example theme() arguments and what aspect of the graph they control: -->

<!-- axis.line : lines forming x-axis and y-axis -->
<!-- axis.line.x: just the line for x-axis -->
<!-- legend.position: positioning of the legend on the graph -->
<!-- panel.background: the background of the graph -->
<!-- panel.border: the border around the graph -->
<!-- title: all titles on the graph -->
```{r theme_axis_line, fig.cap="Using theme argument axis.line to modify x-axis and y-axis lines"}
ggplot(txhousing, aes(x=volume, y=sales, color=listings)) + 
  geom_point() +
  theme(axis.line=element_line(color="black", size=2))
```
<!-- However, the background of the graph, controlled by theme() argument panel.background is a rectangle, so parameters like fill color and border color can be specified element_rect() -->
```{r theme_panel_background, fig.cap="Using theme element axis.line.x to modify x-axis line"}
ggplot(txhousing, aes(x=volume, y=sales, color=listings)) + 
  geom_point() +
  theme(axis.line=element_line(color="black", size=2),
        panel.background=element_rect(fill="white", color="gray"))
```
<!-- With element_text() we can control properties such as the font family or face ("bold", "italic", "bold.italic") of text elements like title, which controls the titles of both axes. -->
```{r theme_title, fig.cap="Using theme argument title to adjust fonts of all titles"}
ggplot(txhousing, aes(x=volume, y=sales, color=listings)) + 
  geom_point() +
  theme(axis.line=element_line(color="black", size=2),
        panel.background=element_rect(fill="white", color="gray"),
        title=element_text(family="serif", face="bold"))
```
#Finally, some theme() arguments do not use element_ functions to control their properties, like legend.position, which simply accepts values "none", "left", "right", "bottom", and "top".
```{r theme_legend_position, fig.cap="Using theme argument legend.position to position legend"}
ggplot(txhousing, aes(x=volume, y=sales, color=listings)) + 
  geom_point() +
  theme(axis.line=element_line(color="black", size=2),
        panel.background=element_rect(fill="white", color="gray"),
        title=element_text(family="serif", face="bold"),
        legend.position="bottom")
```

```{r theme_classic, fig.cap="theme_classic()"}
ggplot(txhousing, aes(x=volume, y=sales, color=listings)) + 
  geom_point() +
  theme_classic()
```

#theme_dark() makes a dramatic change to the look:
```{r theme_dark, fig.cap="theme_dark()"}
ggplot(txhousing, aes(x=volume, y=sales, color=listings)) + 
  geom_point() +
  theme_dark()
```

## Saving Plots

```{r save_plot, eval=FALSE}
# Save last displayed plot as pdf
ggsave("plot.pdf")

# If you're working with lots of graphs, you can store them in R objects
p <- ggplot(Sitka, aes(x=Time, y=size)) +
  geom_point()
# Use the plot argument of ggsave() to specify which plot to save
ggsave("myplot.png", plot=p)
```

## Line Graph Examples

<!-- The birthwt data set contains data regarding risk factors associated with -->
<!-- low infant birth weight. -->
<!-- The data consist of 189 observations of 10 variables, all numeric: -->
<!-- low: 0/1 indicator of birth weight < 2.5 kg -->
<!-- age: mother’s age -->
<!-- lwt: mother’s weight in pounds -->
<!-- race: mother’s race, (1=white, 2=black, 3=other) -->
<!-- smoke: 0/1 indicator of smoking during pregnancy -->
<!-- ptl: number of previous premature labors -->
<!-- ht: 0/1 indicator of history of hypertension -->
<!-- ui: 0/1 indicator of uterine irritability -->
<!-- ftv: numer of physician visits during first trimester -->
<!-- bwt: birth weight -->
<!-- Let’s take a look at the structure of the birthwt data set first, to get an -->
<!-- idea of how the variables are measured. -->

```{r color numeric, fig.cap="color gradient is appropriate for a continuous variable"}
ggplot(birthwt, aes(x=age, y=bwt, color=race)) +
  geom_point()
```

```{r color factor, fig.cap="evenly spaced hues emphasize contrasts between groups of a factor"}
ggplot(birthwt, aes(x=age, y=bwt, color=factor(race))) +
  geom_point()
```

```{r shape factor, fig.cap="evenly spaced hues emphasize contrasts between groups of a factor"}
ggplot(birthwt, aes(x=age, y=bwt, shape=factor(race))) +
  geom_point()
```

```{r size discrete, fig.cap="size and alpha should only be mapped to numeric variables", warning=TRUE}
ggplot(birthwt, aes(x=age, y=bwt, size=factor(race))) +
  geom_point()
```

## Factor Variables

```{r smoke factor}
birthwt$smokef <- factor(birthwt$smoke, levels=0:1, labels=c("did not smoke", "smoked"))
birthwt$racef <- factor(birthwt$race, levels=1:3, labels=c("white", "black", "other"))
```

## Boxplot of Birthweight by Smoking Status

```{r weight by smoke, fig.cap="birth weight distribution by mother's smoking status"}
ggplot(birthwt, aes(y=bwt, fill=smokef)) +
  geom_boxplot()
```

## Overlapping Points

```{r overlapping.points, fig.cap="too many discrete values leads to overlapping points"}
ggplot(birthwt, aes(x=racef, y=age)) +
  geom_point()
```

```{r jitter, fig.cap="jittering adds random variation to the position of the points"}
ggplot(birthwt, aes(x=racef, y=age)) +
  geom_point(position="jitter")
```

## Bar Charts

```{r stack, fig.cap="geom_bar() will stack bars with the same x-position"}
ggplot(birthwt, aes(x=low, fill=racef)) +
  geom_bar()
```

```{r dodge, fig.cap="dodging emphasizes counts"}
ggplot(birthwt, aes(x=low, fill=racef)) +
  geom_bar(position="dodge")
```

```{r fill, fig.cap="filling emphasizes proportions"}
ggplot(birthwt, aes(x=low, fill=racef)) +
  geom_bar(position="fill")
```

